Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-23 LICS_2025/monoidal_e_graphs/sections/introduction.tex
wip

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations and the foundations of e-graphs in particular.
We propose to address this gap via a categorical axiomatisation for e-graphs and their rewriting.
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and morphisms of Cartesian categories enriched over a category of semilattices (\textit{semilattice-enriched}).
The morphisms of the latter can also be depicted using \emph{string diagrams}~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}: a two-dimensional syntax that encodes equivalence classes of morphism terms.
Enriched categories generalise plain categories by imposing more structure making the categorical language more expressive~\cite{Borceux_1994,Kelly2022BASICCO} which will be useful for modelling equivalence between terms as encountered in e-graphs.
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, for which there are existing precedents in the literature~\cite{bonchi_string_2022,bonchi_string_2022-1,bonchi_string_2022-2,ghica_rewriting_2023,ghica_hierarchical_2023}.
The key ingredient of our categorical approach is the semilattice-enrichment.
By the latter we will mean an idempotent commutative semigroup given by a binary operator but will stick to the word semilattice as it is shorter.
By using the binary operator of a given semilattice we can operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

Although our primary interest is theoretical, we believe that phrasing e-graphs in the more general language of category theory will open the door to new potential domains of applications that are already phrased in categorical language with a defined notion of rewriting and thus these applications will benefit from the mentioned optimisation technique.
% now use string diagrams and which require the application of optimisation techniques \cite{Selinger_2010, Hasegawa-traced}.
This may include areas as diverse as digital \cite{ghica_compositional_2023} and quantum circuits
\cite{coecke_interacting_2011,ZX}, functional programs~\cite{ghica-zanassi2023string}, or computational linguistics \cite{wazni_quantum_2022,coecke_lambek_2013}.
% However, in this work we offer little to the e-graph practitioner, as we focus exclusively on foundational aspects, introducing no new applications or improved e-graph algorithmics.

A final contribution of this work is giving a concrete combinatorial representation of string diagrams for semilattice-enriched categories in terms of (hierarchical)  hypergraphs, which we call \textit{e-hypergraphs}.
We also give a specification of e-hypergraph rewriting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{dpo, bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022},  proving the representation \textit{sound and complete} with respect to our categorical semantics.  
As a corollary, this also provides a formalisation of the rewriting theory of e-graphs as encountered in the literature~\cite{EggPaper}.

@@

\ifdefined \ONECOLUMN
\begin{figure}
	\[
		\scalebox{0.55}{
		\tikzfig{../figures/categorical-semantics/egraph-translation}
		}
	\]
	\caption{Example translation of acyclic e-graphs into string diagrams for semilattice-enriched Cartesian categories. }
	\label{fig:e-graph-example}
	\end{figure}
\else
\begin{figure*}
\[
	\hspace{1.3cm}
    \scalebox{0.65}{
    \tikzfig{../figures/categorical-semantics/egraph-translation}
    }
\]
\caption{Example translation of acyclic e-graphs into string diagrams for semilattice-enriched Cartesian categories. }
\label{fig:e-graph-example}
\end{figure*}
\fi

@@

Nodes are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs rooted at these boxes.
Initially, each equivalence class has a single node (subgraph).
The first rewrite rule, replacing multiplication by the more efficient \emph{shift-left} operator, creates the first non-unitary (containing more than one node or subgraph) equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) operator nodes, with a new sharing: that of node $a$.
This second e-graph illustrates the other difference, besides sharing, between conventional e-graphs and e-hypergraphs: in the former, edges connect directly to nodes inside the equivalence class, whereas in the latter, edges connect to the equivalence class itself.
Explicit discarding (weakening) nodes, depicted with black dots with a single wire inside the dashed boxes 
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}, indicate which of the class-level edges are connected to which nodes inside the class.

@@

\begin{figure}
\[
	\scalebox{0.65}{
	\tikzfig{../figures/categorical-semantics/egraph-strings}
	}
\]
\captionsetup{belowskip=-4ex}
\caption{String diagrams for semilattice-enriched symmetric monoidal categories.}
\label{fig:egraph-strings}
\end{figure}

@@

String diagrams are read from bottom-to-top, and we implicitly consider additional generators for the duplication and deletion transformations of a (semilattice-enriched) Cartesian category.
Note, further, that in the Cartesian case we can restrict generating morphisms $c$ to have a single output,  without loss of generality,  by applying the universal properties of the Cartesian product.  
This means the (informal) translation given below is fully general for acyclic e-graphs.

The translation from e-graphs to enriched string diagrams is illustrated informally in Figure~\ref{fig:e-graph-to-string},
noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding, in each component, all unnecessary inputs.

\begin{figure}
\[
    \tikzfig{../figures/categorical-semantics/egraph-translation-1}
\]
\caption{E-graph e-class translation to string diagram.}
\label{fig:e-graph-to-string}
\end{figure}
The subfigure on the right is slightly simplified: if we denote the arity of each $c_i$ as $a_{c_{i}}$, then the dashed box on the right should contain $\sum\limits_{i} a_{c_{i}}$ inputs and each component $c_j$ has $\sum\limits_{i \not = j} a_{c_{i}}$ weakening nodes inside.
Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}.
% In particular, note that the translation of the \textit{cyclic} e-graph (e) requires the use of a categorical \textit{trace}, generating a cycle. 
% In this paper, we focus on \textit{acyclic} e-graphs and their corresponding categories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

The motivating example of our work, e-graphs, requires the Cartesian structure to express arbitrary sharing as used in e-graphs.
However, we will see how all the core of the theory of e-hypergraph rewriting only requires the monoidal structure. 
Thus, we can readily generalise e-graphs from algebraic to monoidal theories, giving rise to a  host of new potential applications, which we briefly outline in the conclusion of this paper.

@@

In order to \textit{implement} generalised e-graphs, rewriting must be performed on concrete representations of the corresponding string diagrams.  
String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo ``connectivity'') or as a 2-dimensional syntax quotiented by the equations of a \textit{symmetric monoidal category (SMC)}.
For example, we have the following equivalences between diagrams:
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/interchange}
	}
\]
The latter representation makes string diagrams unamenable to efficient implementation due to the difficulty of calculating the quotient.
\emph{Hypergraphs} are a combinatorial representation designed to address this issue~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.
% A large body of research work shows how this issue can be solved by taking a different approach, representing string diagrams as hypergraphs~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  
Here, the generators $c_i$ become hyper-edges,  and wires become vertices.  
Thus, the expected quotient becomes simply hypergraph isomorphism. 
With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterise} the free symmetric monoidal category generated by some~$c_i$.

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(d) of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  Because the generating equations can be applied in any context, we are led to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ within a hypergraph and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{dpo, bonchi_string_2022}, is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs} --- hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

@@

% String diagrams
Our use of string diagrams is also central to the intuition behind our work \cite{Selinger_2010, joyal_geometry_1991}.  
There is a breadth of work on similar string diagrammatic formalisms,  including hierarchical string diagrams \cite{ghica-zanassi2023string} and functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}. 
% Hierarchical hypergraphs
Hierarchical (hyper-)graphs are also a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}. 
% String diagram rewrites
Our work builds heavily on string diagram rewrite theory as developed in \cite{bonchi_string_2022,bonchi_string_2022-1, bonchi_string_2022-2}.

@@

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$,  or simply $A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.  
We denote by $\catname{SLat}$ the closed monoidal category of semilattices which is defined in section~\ref{sec:appendix:slat} of the Appendix.

@@

It will often be convenient to give presentations of SMTs in terms of string diagrams.
\begin{example}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
	\scalebox{0.8}{
  	 \tikzfig{../figures/categorical-semantics/Cartesian-equipment}
	}
\]
together with the following associativity, commutativity and unitality equations, given in terms of equations of string diagrams. 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-theory}	
	}
\]
A Cartesian SMT is given by a set of generators $\Sigma_C$ which additionally contains the generators of commutative comonoids, together with the equations of the SMT of commutative comonoids, and the following additional naturality of copy-delete equations, for every $c \in \Sigma_C$
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-naturality}
	}
\]
By Fox's Theorem~\cite{fox},  $\textbf{S}(\Sigma_C, \mathcal{E}_C)$ is equivalent to the free Cartesian category over $\Sigma_C$. 
\end{example}

@@

We will use enrichment to axiomatize the e-box (e-class) structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. 
More precisely our hom-sets will be given the structure of a semilattice.
% Note, in particular,  that we do not require semilattices to have a unit for $+$. 
% This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  
Generally,  we may define a category \textit{enriched} in any other (closed) monoidal category $M$ (\textit{e.g.},  $\catname{SLat}$).
In this case,  hom-sets are generalised to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
\[
	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
\]
is defined as an $M$-morphism which satisfies certain axioms.
Composition being an $M$-morphism implies it respects the additional structure on hom-sets.
%  In other words,  composition must respect the additional structure on hom-sets.  
% In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
We omit further details of the general case of enrichment here, and instead work with the concrete definition required for our application, where $M = \catname{SLat}$.

@@

\begin{proposition}(A specialized case of Proposition 6.4.3~\cite{Borceux_1994}. See also Appendix section~\ref{sec:appendix:slat})
There is a 2-adjunction 
% https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
\[\begin{tikzcd}
	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
	\end{tikzcd}
\]
that is induced by a usual free-forgetful adjunction

\[\begin{tikzcd}
	\catname{SLatt} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
	\end{tikzcd}
\]
\end{proposition}

>>>>

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations and the foundations of e-graphs in particular.
We propose to address this gap via a categorical axiomatisation for e-graphs and their rewriting.
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and morphisms of Cartesian categories enriched over a category of semilattices (\textit{semilattice-enriched}).
% The morphisms of the latter can also be depicted using \emph{string diagrams}~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}: a two-dimensional syntax that encodes equivalence classes of morphism terms.
% Enriched categories~\cite{Borceux_1994,Kelly2022BASICCO} generalise plain categories by imposing more structure making the categorical language more expressive~\cite{Borceux_1994,Kelly2022BASICCO} which will be useful for modelling equivalence between terms as encountered in e-graphs.
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, for which there are existing precedents in the literature~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022-3,fscd,ghica-zanassi2023string}.
The key ingredient of our categorical approach is the semilattice-enrichment.
By the latter we will mean an idempotent commutative semigroup given by a binary operator but will stick to the word semilattice as it is shorter.
By using the binary operator of a given semilattice we can operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

Although our primary interest is theoretical, we believe that phrasing e-graphs in the more general language of category theory will open the door to new potential domains of applications that are induced by monoidal rather than algebraic theories with a defined notion of rewriting and thus these applications will benefit from the mentioned optimisation technique.
% now use string diagrams and which require the application of optimisation techniques \cite{Selinger_2010, Hasegawa-traced}.
This may include areas as diverse as digital \cite{ghica_compositional_2023} and quantum circuits
\cite{coecke_interacting_2011,ZX}, functional programs~\cite{ghica-zanassi2023string}, or computational linguistics \cite{wazni_quantum_2022,coecke_lambek_2013}.
% However, in this work we offer little to the e-graph practitioner, as we focus exclusively on foundational aspects, introducing no new applications or improved e-graph algorithmics.

A final contribution of this work is giving a concrete combinatorial representation of string diagrams for semilattice-enriched categories in terms of (hierarchical)  hypergraphs, which we call \textit{e-hypergraphs} or monoidal e-graphs, as the former are a generalisation of e-graphs for monoidal theories.
We also give a specification of e-hypergraph rewriting via a suitable extension of the double pushout (DPO) rewriting framework 
\cite{dpo, bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022-3},  proving the representation \textit{sound and complete} with respect to our categorical semantics.  
As a corollary, this also provides a formalisation of the rewriting theory of e-graphs as encountered in the literature~\cite{EggPaper}.

@@

\ifdefined \ONECOLUMN
\begin{figure}
	\[
		\scalebox{0.55}{
		\tikzfig{../figures/categorical-semantics/egraph-translation}
		}
	\]
	\caption{Example translation of acyclic e-graphs into string diagrams for semilattice-enriched Cartesian categories. }
	\label{fig:e-graph-example}
	\end{figure}
\else
\begin{figure*}
\[
	\hspace{1.3cm}
    \scalebox{0.65}{
    \tikzfig{../figures/categorical-semantics/egraph-translation}
    }
\]
\captionsetup{skip=0pt, belowskip=-5mm}
\caption{Example translation of acyclic e-graphs into string diagrams for semilattice-enriched Cartesian categories. }
\label{fig:e-graph-example}
\end{figure*}
\fi

@@

Nodes are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs rooted at these boxes.
Initially, each equivalence class has a single node (subgraph).
The first rewrite rule, replacing multiplication by the more efficient \emph{shift-left} operator, creates the first non-unitary (containing more than one node or subgraph) equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) operator nodes, with a new sharing: that of node $a$.
This second e-graph illustrates the other difference, besides sharing, between conventional e-graphs and e-hypergraphs: in the former, edges connect directly to nodes inside the equivalence class, whereas in the latter, edges connect to the equivalence class itself.
Explicit discarding (delete) nodes, depicted with black dots with a single wire inside the dashed boxes 
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}, indicate which of the class-level edges are connected to which nodes inside the class.

@@

\begin{figure}
\[
	\scalebox{0.65}{
	\tikzfig{../figures/categorical-semantics/egraph-strings}
	}
\]
\captionsetup{skip=0pt, belowskip=-4ex}
\caption{String diagrams for semilattice-enriched symmetric monoidal categories.}
\label{fig:egraph-strings}
\end{figure}

@@

String diagrams are read from bottom-to-top, and we implicitly consider additional generators for the duplication and deletion transformations of a (semilattice-enriched) Cartesian category.
Note, further, that in the Cartesian case we can restrict generating morphisms $c$ to have a single output,  without loss of generality,  by applying the universal properties of the Cartesian product.  
This means the (informal) translation given next is fully general for acyclic e-graphs.

The translation from e-graphs to enriched string diagrams is illustrated informally in Figure~\ref{fig:e-graph-to-string},
noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding, in each component, all unnecessary inputs.
Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}.

\begin{figure}
\[
    \tikzfig{../figures/categorical-semantics/egraph-translation-1}
\]
\captionsetup{skip=0pt}
\caption{E-graph e-class translation to string diagram.}
\label{fig:e-graph-to-string}
\vspace{-8mm}
\end{figure}
% The subfigure on the right is slightly simplified: if we denote the arity of each $c_i$ as $a_{c_{i}}$, then the dashed box on the right should contain $\sum\limits_{i} a_{c_{i}}$ inputs and each component $c_j$ has $\sum\limits_{i \not = j} a_{c_{i}}$ weakening nodes inside.
% Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}.
% In particular, note that the translation of the \textit{cyclic} e-graph (e) requires the use of a categorical \textit{trace}, generating a cycle. 
% In this paper, we focus on \textit{acyclic} e-graphs and their corresponding categories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

The motivating example of our work, e-graphs, requires the Cartesian structure to express arbitrary sharing as could be seen in Figure~\ref{fig:e-graph-example}.
However, we will see how all the core of the theory of e-hypergraph rewriting only requires the monoidal structure. 
Thus, we can readily generalise e-graphs from algebraic to monoidal theories, giving rise to a  host of new potential applications, which we briefly outline in the conclusion of this paper.

@@

In order to \textit{implement} generalised e-graphs, rewriting must be performed on concrete representations of the corresponding string diagrams.  
String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo ``connectivity'') or as a 2-dimensional syntax quotiented by the equations of a \textit{symmetric monoidal category (SMC)}.
For example, we have the following equivalences between diagrams:
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/interchange}
	}
\]
The latter representation makes string diagrams unamenable to efficient implementation due to the difficulty of calculating the quotient.
\emph{Hypergraphs} are a combinatorial representation designed to address this issue~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022-3}.
% A large body of research work shows how this issue can be solved by taking a different approach, representing string diagrams as hypergraphs~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  
Here, the generators $c_i$ become hyper-edges,  and wires become vertices.  
Thus, the expected quotient becomes simply hypergraph isomorphism. 
With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterise} the free symmetric monoidal category generated by some~$c_i$.

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(d) of Figure \ref{fig:e-graph-example}. 
These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  
Because the generating equations can be applied in any context, we are led to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ within a hypergraph and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{dpo, bonchi_string_2022-1}, is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs} --- hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

@@

% String diagrams
Our use of string diagrams is also central to the intuition behind our work \cite{Selinger_2010, joyal_geometry_1991}.  
There is a breadth of work on similar string diagrammatic formalisms,  including hierarchical string diagrams \cite{ghica-zanassi2023string} and functorial boxes \cite{mellies_functorial_2006},  proof nets for compact closed categories with biproducts \cite{duncan_generalised_2009}, and recent work on tape diagrams for rig categories with biproducts \cite{bonchi_tape_nodate}. 
% Hierarchical hypergraphs
Hierarchical (hyper-)graphs are also a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}. 
% String diagram rewrites
Our work builds heavily on string diagram rewrite theory as developed in \cite{bonchi_string_2022-1,bonchi_string_2022-2, bonchi_string_2022-3}.

@@

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.  
We denote by $\catname{SLat}$ the closed monoidal category of semilattices which is defined in section~\ref{sec:appendix:slat} of the Appendix.
Given two adjoint functors $F : \mathbb{A} \to \mathbb{B}$ and $G : \mathbb{B} \to \mathbb{A}$ with $F$ being left-adjoint, we write $F \dashv G : \mathbb{B} \to \mathbb{A}$ for this adjunction.

@@

It will often be convenient to give presentations of SMTs in terms of string diagrams.
\begin{example}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
	\scalebox{0.8}{
  	 \tikzfig{../figures/categorical-semantics/Cartesian-equipment}
	}
\]
together with the following associativity, commutativity and unitality equations, given in terms of equations of string diagrams. 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-theory}	
	}
\]
A Cartesian SMT is given by a set of generators $\Sigma_C$ which additionally contains the generators of commutative comonoids, together with the equations $\mathcal{E}$ that include the SMT of commutative comonoids, and the following additional naturality of copy-delete equations, for every $c \in \Sigma_C$
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-naturality}
	}
\]
By Fox's Theorem~\cite{fox},  $\textbf{S}(\Sigma_C, \mathcal{E}_C)$ is equivalent to the free Cartesian category over $\Sigma_C$. 
\end{example}

@@

We will use enrichment to axiomatize the e-box (e-class) structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. 
More precisely our hom-sets will be given the structure of a semilattice.
% Note, in particular,  that we do not require semilattices to have a unit for $+$. 
% This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  
Generally,  we may define a category \textit{enriched} in any other (closed) monoidal category $M$, but we will omit further details of the general case of enrichment here, and instead work with the concrete definition required for our application, where $M = \catname{SLat}$.
%  (\textit{e.g.},  $\catname{SLat}$).
% In this case,  hom-sets are generalised to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
% \[
% 	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
% \]
% is defined as an $M$-morphism which satisfies certain axioms.
% Composition being an $M$-morphism implies it respects the additional structure on hom-sets.
%  In other words,  composition must respect the additional structure on hom-sets.  
% In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
% We omit further details of the general case of enrichment here, and instead work with the concrete definition required for our application, where $M = \catname{SLat}$.

@@

\begin{proposition}(A specialized case of Proposition 6.4.3~\cite{Borceux_1994}. See also Appendix section~\ref{sec:appendix:slat})
There is a 2-adjunction 
% https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
% \[\begin{tikzcd}
% 	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
% 	\end{tikzcd}
% \]
\vspace{-1mm}
\[
\mathcal{F} \dashv \mathcal{U} : \catname{SLat}\text{--}\catname{Cat} \to \catname{Cat}
\vspace{-1mm}
\]
that is induced by a usual free-forgetful adjunction

% \[\begin{tikzcd}
% 	\catname{SLatt} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
% 	\end{tikzcd}
% \]
\vspace{-1mm}
\[
F \dashv U : \catname{SLat} \to \catname{Set}
\vspace{-1mm}
\]
\end{proposition}
In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free $\catname{SLat}$-category $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
% SMC $(\mathbb{C}, \otimes, 1,+)$ where every hom-set additionally has the structure of a semilattice $(\mathbb{C}(A,B), +)$ -- that is,  a set equipped with an associative,  commutative,  and idempotent operation -- which respects the composition and tensor product in the following ways,  for all appropriately typed morphisms.
% \begin{align*}
% f ; (g+h) &= f;g + f;h &
% (f+g) ; h &= f;h + g;h \\
% f \otimes (g+h) &= f \otimes g + f \otimes h & 
% (f+g) \otimes h &= f \otimes h + g \otimes h,
% %\item A set $ob(\mathbb{C})$ of objects;
% %\item For every pair $A,B \in ob(\mathbb{C})$, a hom-semigroup $Hom(A,B)$;
% %\item An element $\textsf{id}_{A,B} \in Hom(A,B)$; 
% \end{align*}
% Note,  we take $(;\!)$ and $\otimes$ to bind more tightly than $+$.
% A \textit{semilattice-enriched strict SMC functor} $F: (\mathbb{C}, \otimes, 1,+) \to (\mathbb{D}, \otimes, 1,+)$ is given by a strict SMC functor on the underlying SMCs which additionally satisfies $F(f+g) = F(f)+F(g)$.
% All definitions extend to \textit{PROP} when the manipulated categories are PROPs rather than SMCs. 
% We denote the \textit{freely enriched PROP (SMT)} over a monoidal signature $\Sigma$ as $\textbf{PROP}^+(\Sigma)$ ($\textbf{SMT}^+(\Sigma, \mathcal{E})$,  respectively). 
% \end{definition}
\begin{definition}
We define $\catname{S}(\Sigma)^{+}$ and $\catname{S}(\Sigma, \mathcal{E})^{+}$ to be $\catname{SLat}$-SMCs obtained as $\mathcal{F}(\catname{S}(\Sigma))$ and $\mathcal{F}(\catname{S}(\Sigma, \mathcal{E}))$ respectively.
Both of these $\catname{SLat}$-categories are in $\catname{PROP}^{+}$ which we define to be the image of $\catname{PROP}$ via the same adjunction.
\end{definition}
Equivalently, they can be constructed from taking (typed) $\Sigma^+$-terms  -- namely, those freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sym_{1,1}$, $(;\!)$ and $\otimes$ (and $f+g$, respectively) -- quotiented by the axioms of an enriched symmetric monoidal category (and the least congruence including $\mathcal{E}$ in the case of $\catname{S}(\Sigma, \mathcal{E})^{+}$).
% The free category $\textbf{PROP}^+(\Sigma)$ can be syntactically constructed from taking (typed) $\Sigma^+$-terms  -- namely, those freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sym{1,1}$, $(;\!)$ and $\otimes$ (and $f+g$, respectively) -- quotiented by the axioms of an enriched symmetric monoidal category.
